### 2019-05-25 reports/thesis/chapters/3_compoundPoisson.tex

>>>>

\section{Compound Poisson}

The greyvalues of each pixel in the detector can be modelled as a random variable, due to the random behaviour of photons being produced, interacting with the phantom and interacting with the scintillator in the detector. When the photons interact with the scintillator, they are converted into visible light. The visible light photons are then detected and converted into a greyvalue. How the detected visible light is converted is important in modelling the greyvalue.

Should the greyvalue be proportional to the number of photons detected, the detector is known as a quantum counter and the greyvalue is a Poisson random variable, up to a constant \citep{whiting2006properties}. However if the greyvalue records the signal strength, proportional to the energy detected, then the detector is an energy integrating detector. The greyvalue has the compound Poisson distribution, again up to a constant \citep{whiting2006properties}. There are other types of detection schemes \citep{whiting2006properties} but it shall be not be considered here.

Experiments have been done to verify the compound Poisson nature of the detector. This was done by investigating the variance of radiographs of air \citep{hsieh2015compound} and a polyethylene cylinder \citep{yang2009evaluation} \citep{yang2010noise} at different voltages and powers. It was found there were 2 components in the noise, one was signal dependent and comes from the compound Poisson, the other was signal independent and may be electronic noise. The electronic noise can be modelled as Normally distributed \citep{xu2009electronic}.

The compound Poisson can be formally written. Let $Y$ be the number of photons detected in a pixel for some time exposure $\tau$,
\begin{equation*}
	Y\sim\poisson(\lambda)
\end{equation*}
that is
\begin{equation*}
	\prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!} \quad \text{for }y=0,1,2,\dotdotdot
\end{equation*}
where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be the energy of the $i$th photon in a pixel for i.i.d.~$i=1,2,3,\dotdotdot$ with p.d.f.~$p_U(u)$. Let $X$ be the greyvalue of a pixel and is an energy integrating detector, then
\begin{equation}
	X|Y = \sum_{i=1}^{Y}U_i+\kappa
	\label{eq:compoundPoisson_X|Y}
\end{equation}
where $\kappa$ is some offset. For now let $\kappa=0$, then the marginal $X$ has the compound Poisson distribution. The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation*}
	p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0
	\ .
\end{equation*}
However it should be noted that $X=0$ if and only if $Y=0$ with probability
\begin{equation*}
	\prob(X=0)=\euler^{-\lambda} \ .
\end{equation*}
Formally $X$ has probability mass at $X=0$ and probability density at $X>0$, then 
\begin{equation}
	p_X(x) = 
	\begin{cases}
		\delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\ 
		\sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
	\end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function. It is quite often the case this cannot be written in closed form.

The m.g.f.~of $X$ has a nicer form however. Let the m.g.f.~of $X$ be
\begin{equation*}
	M_X(\theta)=\expectation\left[\euler^{X\theta}\right]
	\ .
\end{equation*}
This can be computed using the result for conditional expectations
\begin{equation*}
	M_X(\theta)=\expectation\expectation\left[\euler^{X\theta}|Y\right]
	\ .
\end{equation*}
Using the definition of $X|Y$ in Equation \eqref{eq:compoundPoisson_X|Y}, then
\begin{equation*}
	M_X(\theta)=\expectation\expectation\left[\exp\left(\theta U_1 + \theta U_2 + \dotdotdot + \theta U_Y\right)|Y\right]
\end{equation*}
\begin{equation*}
	M_X(\theta)=\expectation\expectation\left[\euler^{\theta U_1}\cdot\euler^{\theta U_2}\cdot\dotdotdot\cdot\euler^{\theta U_Y}|Y\right]
	\ .
\end{equation*}
But because $U_i$ for $i=1,2,3,\dotdotdot$ are i.i.d., then each $U_i$ has a common m.g.f.
\begin{equation*}
	M_U(\theta)=\expectation\left[\euler^{U\theta}\right]
	\ ,
\end{equation*}
then
\begin{equation*}
	M_X(\theta)=\expectation\left(
		\expectation\left[\euler^{\theta U_1}|Y\right]\cdot
		\expectation\left[\euler^{\theta U_2}|Y\right]\cdot
		\dotdotdot \cdot
		\expectation\left[\euler^{\theta U_Y}|Y\right]
	\right)
\end{equation*}
\begin{equation*}
	M_X(\theta)=\expectation\left[\left(M_U(\theta)\right)^Y\right]
\end{equation*}
\begin{equation*}
	M_X(\theta)=\expectation\left[\euler^{Y\ln(M_U(\theta))}\right]
\end{equation*}
\begin{equation*}
	M_X(\theta) = M_Y\left(\ln(M_U(\theta)\right)
\end{equation*}
where $M_Y(\theta)$ is the m.g.f.~of $Y$. It can be shown that the m.g.f.~of $Y$ is
\begin{equation}
	M_Y(\theta)=\expectation\left[\euler^{Y\theta}\right]=
	\exp
	\left[
		\lambda
		\left(
		  \euler^\theta-1
		\right)
	\right]
\end{equation}
then
\begin{equation}
	M_X(\theta)=
	\exp\left[
		\lambda
		\left(
			M_U(\theta)-1
		\right)
	\right]
	\ .
\end{equation}
As long as the m.g.f.~of $U$ is known, then m.g.f.~of $X$ can be written in closed form.

Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting it to zero
\begin{align*}
	M_X'(\theta)&=\exp\left[\lambda\left(M_U(\theta)-1\right)\right]\cdot\lambda M_U'(\theta) \\
	M_X'(\theta)&=M_X(\theta)\lambda M_U'(\theta)
\end{align*}
then
\begin{equation}
	\expectation\left[X\right]=\lambda\expectation\left[U\right]
	\ .
\end{equation}
Conducting the same procedure
\begin{equation*}
	M_X''(\theta)=M_X'(\theta)\lambda M_U'(\theta)+M_X(\theta)\lambda M_U''(\theta)
	\ ,
\end{equation*}
the variance can be obtained
\begin{equation*}
	\variance\left[X\right]=M_X''(0)-\left[M_X'(0)\right]^2
\end{equation*}
\begin{equation*}
	\variance\left[X\right]=M_X'(0)\lambda M_U'(0)+M_X(0)\lambda M_U''(0)-\left[\lambda\expectation\left[U\right]\right]^2
\end{equation*}
\begin{equation*}
	\variance\left[X\right]=\lambda^2 \left(\expectation\left[U\right]\right)^2+\lambda \expectation\left[U^2\right]-\left[\lambda\expectation\left[U\right]\right]^2
\end{equation*}
resulting in
\begin{equation}
	\variance\left[X\right] = \lambda\expectation\left[U^2\right]
	\ .
\end{equation}
This can be extended for higher moments.

Because the m.g.f.~can be written down in closed form, so is the characterisation function or Fourier transform of the p.d.f.~of $X$. By obtaining an empirical version of $p_U(u)$, moments of $X$ can be estimated using the m.g.f.~and the p.d.f.~can be estimated by using fast Fourier transform on the empirical characteristic function \citep{whiting2006properties}.

If the distribution of $U$ cannot be obtained or estimated, then one can guess that distribution. \cite{xu2009electronic} for example used a Gamma distribution
\begin{equation}
	U\sim\gammaDist\left(\alpha,\beta\right)
\end{equation}
where $\alpha>0$ is the Gamma shape parameter and $\beta>0$ is the Gamma rate parameter. This is a special case of the compound Poisson called the compound Poisson-Gamma model and one can write
\begin{equation}
	X\sim\CPoisson(\lambda,\alpha,\beta)
	\ .
\end{equation}
The conditional distribution can be shown to be
\begin{equation}
	X|Y\sim\gammaDist\left(Y\alpha,\beta\right)
	\ .
\end{equation}

Using results derived earlier, the m.g.f.~of the of $X$ is
\begin{equation}
	M_X(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right]
\end{equation}
and moments can be obtained from it such as
\begin{equation}
	\expectation\left[X\right]=\frac{\alpha\lambda}{\beta}
\end{equation}
\begin{equation}
	\variance\left[X\right]=\frac{\alpha(\alpha+1)\lambda}{\beta^2}
	\label{eq:compoundPoisson_variance}
\end{equation}
and
\begin{equation}
	\expectation\left[(X-\expectation[X])^3\right] = \frac{\alpha(\alpha+1)(\alpha+2)\lambda}{\beta^3}
	\ .
\end{equation}

The p.d.f.~can be shown to be
\begin{equation}
	p_X(x) = 
	\begin{cases}
		\delta(x) \euler^{-\lambda} & \text{ for } x=0 \\ 
		\sum_{y=1}^{\infty}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!} & \text{ for } x>0
	\end{cases}
	\label{eq:compoundPoisson_pdf}
\end{equation}
and it is well known the infinite sum cannot be written in closed form. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Of course direct simulation of a compound Poisson-Gamma random variable is incredibly easy if $U$ can be simulated. The p.d.f.~then can be estimated using these simulated samples.

For a given i.i.d.~random sample of a compound Poisson-Gamma random variable, parameter estimation can be difficult. Maximum likelihood estimation is very complicated \citep{withers2011compound}. It can be made simpler by fixing $\alpha$ to be known \citep{withers2011compound} but this is not useful in this context.

Because the moments of a compound Poisson-Gamma random variable can be obtained easily, method of moments estimators can be obtained. Suppose $\widehat{\mu}$ is an estimator of $\expectation[X]$ and $\widehat{\mu}_j$ is an estimator of $\expectation\left[\left(X-\expectation[X]\right)^j\right]$ for $j=2,3$. Then the estimators
\begin{equation}
	\widehat{\lambda}=\frac{\widehat{\mu}^2\widehat{\mu}_2}{\left(2\widehat{\mu}_2^2-\widehat{\mu}_3\widehat{\mu}\right)}
\end{equation}
\begin{equation}
	\widehat{\alpha}=\frac{\widehat{\mu}_3\widehat{\mu}-2\widehat{\mu}_2^2}{\widehat{\mu}_2^2-\widehat{\mu}\widehat{\mu}_3}
\end{equation}
\begin{equation}
	\widehat{\beta}=\frac{\widehat{\mu}\widehat{\mu}_2}{\widehat{\mu}\widehat{\mu}_3-\widehat{\mu}_2^2}
\end{equation}
are method of moments estimators of $\lambda$, $\alpha$ and $\beta$ respectively \citep{withers2011compound}. These estimators suffer because estimation is not done through the sufficient statistics and can be negative, this is a problem because the parameters do not take non-positive values.

However, parameter estimation can be made easier once it can be realised that the compound Poisson-Gamma distribution is in the exponential family for fixed $\alpha$ \citep{jorgensen1987exponential}. To show this, it is particular useful to reparametrize the compound Poisson-Gamma distribution using the following:
\begin{equation}
	p=\frac{2+\alpha}{1+\alpha}
	\ ,
\end{equation}
\begin{equation}
	\mu=\frac{\lambda\alpha}{\beta}
	\ ,
\end{equation}
\begin{equation}
	\phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}}
	\ .
\end{equation}
The parameters $p$, $\mu$ and $\phi$ are called the index, mean and dispersion parameters respectively. It can be shown that $1<p<2$, $\mu>0$ and $\phi>0$.

By rearranging the parameters to get
\begin{equation}
	\lambda=\frac{\mu^{2-p}}{\phi(2-p)}
\end{equation}
\begin{equation}
	\alpha=\frac{2-p}{p-1}
\end{equation}
\begin{equation}
	beta=\frac{1}{\phi(p-1)\mu^{p-1}}
\end{equation}
and substituting it into Equation \eqref{eq:compoundPoisson_pdf}, the p.m.f.~at zero can be shown to be
\begin{equation}
	\prob(X=0) = \exp
	\left[
	    -\frac{\mu^{2-p}}{\phi(2-p)}
	\right]
\end{equation}
and the p.d.f.~for $x>0$ is
\begin{multline*}
	p_X(x) = \sum_{y=1}^{\infty}
	\left[
		\frac{1}{\phi(p-1)\mu^{p-1}}
	\right]^{y\alpha}
	\frac{1}{\Gamma(y\alpha)}
	x^{y\alpha-1}
	\exp\left[
	    -\frac{x}{\phi(p-1)\mu^{p-1}}
	\right]
	\\
	\exp\left[
	    -\frac{\mu^{2-p}}{\phi(2-p)}
	\right]
	\left[
		\frac{\mu^{2-p}}{\phi(2-p)}
	\right]^y
	\frac{1}{y!}
	\ .
\end{multline*}
Tidying up the equation
\begin{multline*}
	p_X(x) = 
	\exp\left[
		\frac{1}{\phi}\left(x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}\right)
	\right]
	\frac{1}{x}
	\\
	\sum_{y=1}^{\infty}\frac{x^{y\alpha}\mu^{y[2-p-\alpha(p-1)]}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
	\ .
\end{multline*}
To simplify further, it should be noted that
\begin{align*}
	2-p-\alpha(p-1) &= 2-p - \frac{2-p}{p-1}(p-1)
	\\&=0
\end{align*}
so that
\begin{equation}
	p_X(x) = 
	\exp\left[
		\frac{1}{\phi}
		\left(
			x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}
		\right)
	\right]
	\frac{1}{x}
	\sum_{y=1}^{\infty}W_y(x,p,\phi)
\end{equation}
where
\begin{equation}
	W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
	\ .
\end{equation}

Those familiar with generalised linear models \citep{nelder1972generalized} \citep{mccullagh1984generalized} will notice that for fixed $p$ or $\alpha$, the above is in the form of a distribution in the dispersive exponential family. Parameter estimation then can be done via the generalised linear model framework and can be extended to include linear mixed models \citep{zhang2013likelihood}. These has applications in for example insurance claim data \citep{jorgensen1994fitting} \citep{smyth2002fitting}.

Estimating $p$ is difficult and various methods were discussed by \cite{zhang2013likelihood}. One easy way is to estimate $\mu$ and $\phi$ on a grid of $p$'s and then select the $p$ which maximises the likelihood \citep{dunn2005series}.

One special property of the compound Poisson-Gamma distribution is that it is in the Tweedie dispersion exponential family \citep{jorgensen1987exponential}. It has a special variance mean relationship
\begin{equation}
	\variance[X] = \phi \mu^p
\end{equation}
where, as a reminder, $1<p<2$. This can be derived using Equation \eqref{eq:compoundPoisson_variance} or via the partition function $Z$. That is let
\begin{equation}
	\theta = \frac{\mu^{1-p}}{1-p}
\end{equation}
and
\begin{equation}
	\ln Z = \frac{1}{\phi(2-p)\theta^\alpha(1-p)^\alpha} \ ,
\end{equation}
then
\begin{equation}
	\variance[X] = \phi^2 \frac{\partial^2\ln Z}{\partial\theta^2} \ .
\end{equation}
